When two-point correlation function fails to describe a structure...

### 2024-03-05 paper.tex
Improve the main result

Now we get our main result:
\begin{theorem}
  Let $A$ be a 2-dimensional $D_1 \times D_2$ array of 0's and 1's where each
  element is chosen randomly and independently with probability $1/2$ with
  exception of three elements $A[0, 0]$, $A[D_1 - 1, 0]$ and $A[0, D_2 - 1]$,
  which are 1's. Then $A$ does not have non-trivial collisions in the two-point
  correlation function with probability $p \to 1$ as $D_1, D_2 \to \infty$.
\end{theorem}
\begin{proof}
  Let $f(x, y)$ be an image of $A$ under Z-transform and a change of variables
  $x = z_1^{-1}, y = z_2^{-1}$. Let $\phi f = f \mod y$ be a
  ``partial evaluation'' homomorphism from $\mathbb{Z}[x, y]$ to
  $\mathbb{Z}[x]$. Let $f_x(x) = \phi f = f(x,0)$. Under conditions of the
  theorem $f(0) = 1$ and $d = \deg f_x = D_2 - 1$.  According to
  \cref{eq:prob-irr} we have $P\{\text{$f_x$ is irreducible}\} \to 1$ with
  $d \to \infty$.

  We have three possible cases of factorization of $f$:
  \begin{itemize}
  \item $f$ is irreducible.
  \item $f$ factors as $f(x, y) = g(x)h(y)$. In this case
    $\phi h = h(0) = const$ and \cref{lem:factors} cannot be applied. According
    to distribution of coefficients in $f$ the probability of this case is
    $P_{split} = 2^{D_1 + D_2 - D_1 D_2 - 1}$.
  \item $f$ factors as $f(x, y) = g(x, y) h(x, y)$. According to conditions on
    the coefficients $\deg_x g > 0$ and $\deg_x h > 0$, therefore
    $\phi g \ne const$ and $\phi h \ne const$. We have
    $f_x = (\phi g) (\phi h)$. As we have shown, the probability of this event
    tends to zero with $D_2 \to \infty$. Applying \cref{lem:factors}, we can say
    the same about probability of this case.
  \end{itemize}
  Hence, the probability of the first case tends to one, with increase in
  $D_1, D_2$. If $f$ is irreducible, there is no two or more non-palindrome
  factors of $f$, hence $A$ does not have non-trivial collisions.
\end{proof}
This theorem also can be proved for multidimensional arrays by recursively
applying the same argument.

>>>>

Now we get our main result:
\begin{theorem}
  Let $A$ be a 2-dimensional $D_1 \times D_2$ array of 0's and 1's where each
  element is chosen randomly and independently with probability $1/2$ with
  exception of two elements $A[0, 0]$ and $A[D_1 - 1, 0]$, which are 1's. Then
  $A$ does not have non-trivial collisions in the two-point correlation function
  with probability at least
  \begin{equation}
    p = 1 - \sqrt{\frac{2}{\pi (D_1 - 1)}}
    \label{eq:main-result}
  \end{equation}
\end{theorem}
\begin{proof}
  Let $f(x, y)$ be an image of $A$ under Z-transform and a change of variables
  $x = z_1^{-1}, y = z_2^{-1}$ with $\deg_x f = d$, where $d = D_1 - 1$. Let
  $\phi f = f \mod y$ be a ``partial evaluation'' homomorphism from
  $\mathbb{Z}[x, y]$ to $\mathbb{Z}[x]$. Let $f_x(x) = \phi f = f(x,0)$. Under
  conditions of the theorem $f(0) = 1$ and $\deg f_x = d$.  According to
  \cref{eq:prob-irr} we have
  \begin{equation*}
    p\{\text{$f_x$ is irreducible}\} = 1 - \sqrt{\frac{2}{\pi d}} + O(d^{-1})
  \end{equation*}
  Now we need to prove a similar result for $f$.

  We have three possible cases of factorization of $f$:
  \begin{itemize}
  \item $f$ is irreducible.
  \item $f$ factors as $f(x, y) = g(x)h(y)$. In this case
    $\phi h = h(0) = const$ and \cref{lem:factors} cannot be applied. According
    to distribution of coefficients in $f$ the probability of this case is
    $p_{split} = 2^{D_1 + D_2 - D_1 D_2 - 1}$. For any fixed integer $D_2 > 1$
    we have $p_{split} = o(D_1^{-1})$. For $D_2 = 1$ we have $f = f_x$ and
    \cref{eq:prob-irr} is applicable directly to $f$.
  \item $f$ factors as $f(x, y) = g(x, y) h(x, y)$. According to conditions on
    the coefficients $\deg_x g > 0$ and $\deg_x h > 0$, therefore
    $\phi g \ne const$ and $\phi h \ne const$. We have
    $f_x = (\phi g) (\phi h)$. Applying \cref{lem:factors}, we can say that
    $f$ is irreducible with probability at least \cref{eq:main-result}.
  \end{itemize}
  Hence $f$ is irreducible with probability at least \cref{eq:main-result}. If
  $f$ is irreducible, there is no two or more non-palindrome factors of $f$,
  hence $A$ does not have non-trivial collisions.
\end{proof}
Surely, $D_1$ can be replaced with $D_2$ in \cref{eq:main-result} if
$A[0, D_2 - 1] = 1$. This theorem also can be proved for multidimensional arrays
by recursively applying the same argument.
